---
title: "Dimension Reduction Lab"
author: "Matthew L Bendall"
date: "July 16, 2015"
output: html_document
---

# 2015 SISBID Module 4 - Unsupervised Learning
##### Genevera I. Allen & Yufeng Liu
### Dimension Reduction Lab

## Data set - Digits Data.
Either use all digits or choose 2-3 digits if computational speed is a problem. Looking at 3's, 8's and 5's are interesting. Note that NMF takes quite a while to run, so you may want to limit the digits considered for that problem.

```{r}
rm(list=ls())
setwd('~/Projects/UW_institute/SISG16/SISBID-Module4.git')
load('UnsupL.Rdata')
filt_digits <- digits[rownames(digits) %in% c(3,5,8),]
```

### Take a look at the data
```{r}
par(mfrow=c(4,4),mar=c(1,1,1,1))
for(i in 1:16){
  imagedigit(digits[i,])
}
```


## Problem 1 - PCA

#### Problem 1a - Apply PCA to this data.
```{r}
digits.svd <- svd(digits)
U <- digits.svd$u
V <- digits.svd$v  #PC loadings
D <- digits.svd$d
Z <- digits%*%V    #PCs
```

###### PC scatterplot
```{r}
z.df <- data.frame(Z)
names(z.df) <- paste0('PC',1:ncol(z.df))
z.df$digit <- rownames(Z)

ggplot(z.df, aes(PC1,PC2,colour=digit)) + geom_point(alpha=0.5)
ggplot(z.df, aes(PC2,PC3,colour=digit)) + geom_point(alpha=0.5)
ggplot(z.df, aes(PC3,PC4,colour=digit)) + geom_point(alpha=0.5)
```

##### PC loadings
```{r}
par(mfrow=c(4,4),mar=c(1,1,1,1))
for(i in 1:16){
  imagedigit(V[,i])
}
```

#### Problem 1b
Do the first several PCs well separate different digits?  Why or why not?

No.

#### Problem 1c 
Use the first several PCs and PC loadings to evaluate the major patterns in the digits data.  Can you come up with a description of the pattern found by each of the first five PCs?


#### Problem 1d
How many PCs are needed to explain 95% of the variance?  You must decide how many PCs to retain.  Which do you pick and why?

###### Variance Explained
```{r}
var.df <- data.frame(idx=1:256)
denom <- sum(D^2)

var.df$varex <- D^2/denom
var.df$cumvar <- cumsum(var.df$varex)

num_pcs <- max(which(var.df$cumvar<0.95))+1

require(gridExtra)
p1 <- ggplot(var.df,aes(idx,varex)) + geom_line()
p2 <- ggplot(var.df,aes(idx,cumvar)) + geom_line()
grid.arrange(p1,p2,ncol=2)
```

## Problem 2 - NMF.

#### Problem 2a
Apply NMF to this data.

```{r}
require("NMF")
# filt_digits <- digits[sample(nrow(digits),930,replace=F),]
filt_digits <- digits[rownames(digits) %in% c(3,8),]

K <- 20
nmffit <- nmf(filt_digits+1, rank=K) #note - this takes a while
W <- basis(nmffit)
H <- coef(nmffit)

#plot archetypes
par(mfrow=c(4,5), mar=c(1,1,1,1))
for(i in 1:K){
  imagedigit(H[i,])
}
#plot scaterrplots of W for interesting archetypes
i = 1; j = 2;
par(mfrow=c(1,1))
plot(W[,i],W[,j],type="n")
text(W[,i],W[,j],rownames(filt_digits),col=rownames(filt_digits),cex=.7)

```


####Problem 2b
Which value of K did you use?  Why?  What happens when you slightly change your chosen K?

#### Problem 2c
Interpret the archetypes found.  Do any of them accurately reflect the different digits?  Which ones?

#### Problem 2d
Plot NMF basis scatterplots of the factors associated with differences between the digits from 2c.  Do these scatterplots well separate the different digits?  Why or why not?


## Problem 3 - ICA.


#### Problem 3a
Apply ICA to this data set.

#### Problem 3b
Which value of K did you use?  Why?  What happens when you slightly change your chosen K?

#### Problem 3c
Interpret the independent image signals found.  Do any other them accurately reflect the different digits?  Which ones?

##Problem 4 - Comparisons.


#### Problem 4a
Compare and contrast PCA, NMF, and ICA on this data set.  Which one best separates the different digits?  Which one reveals the most interesting patterns?

#### Problem 4b
Overall, which method do you recommend for this data set and why?


##########
#Additional Data set - NCI Microarray data
#(If you have time - take a further look at this data set using various methods for dimension reduction.  Also you may be interested in trying MDS to visualize this data.)
###########



###############################################################
###############################################################
#R scripts to help out with the Dimension Reduction Lab
#Don't peek at this if you want to practice coding on your own!!
##################################################################
#code for digits - ALL
load("UnsupL.Rdata")

#visulaize
par(mfrow=c(4,8))
for(i in 1:32){
  imagedigit(digits[i,])
}

########Problem 1 - PCA
#PCA - take SVD to get solution
#don't center and scale to retain interpretation as images
svdd = svd(digits)
U = svdd$u
V = svdd$v #PC loadings
D = svdd$d
Z = digits%*%V #PCs

#PC scatterplot
i = 4; j = 3;
plot(U[,i],U[,j],type="n")
text(U[,i],U[,j],rownames(digits),col=rownames(digits),cex=.7)

#PC loadings
par(mfrow=c(3,5))
for(i in 1:15){
  imagedigit(V[,i])
}

#Variance Explained
varex = 0
cumvar = 0
denom = sum(D^2)
for(i in 1:256){
  varex[i] = D[i]^2/denom
  cumvar[i] = sum(D[1:i]^2)/denom
}

#screeplot
par(mfrow=c(1,2))
plot(1:256,varex,type="l",lwd=2,xlab="PC",ylab="% Variance Explained")
plot(1:256,cumvar,type="l",lwd=2,xlab="PC",ylab="Cummulative Variance Explained")


#########Problem 2 - NMF
#NMF

require("NMF")

dat38 = rbind(digits[which(rownames(digits)==3),],digits[which(rownames(digits)==8),])

K = 20
nmffit = nmf(dat38+1,rank=K) #note - this takes a while
W = basis(nmffit)
H = coef(nmffit)

#plot archetypes
par(mfrow=c(4,5))
for(i in 1:K){
  imagedigit(H[i,])
}

#plot scaterrplots of W for interesting archetypes
i = 11; j = 15;
par(mfrow=c(1,1))
plot(W[,i],W[,j],type="n")
text(W[,i],W[,j],rownames(dat38),col=rownames(dat38),cex=.7)

##################
#Problem 3 - ICA

K = 20
icafit = fastICA(t(digits),n.comp=K)

#plot independent source signals 
par(mfrow=c(4,5))
for(i in 1:K){
  imagedigit(icafit$S[,i])
}

#################################################################
